A Constraint Solving Approach to Parikh Images of Regular Languages

### 2023-07-11 example.tex
refined description

We will now proceed to show how we prove this using the three principles of
\Calculus{}: problem-aware case splitting (sometimes called branching), lazy
enforcement of connectivity constraints (essentially consisting of a constraint
programming-style propagator), and lazy computation of products of automata.

Both lazinesses are in contrast to the eager approach to finding the Parikh
image described in~\cite{generate-parikh-image}, that would have started by
computing the product $A' \times B$, translated it to a Presburger formula with
$i, n$ etc as free variables, and then plugged in
\cref{const:more-inside-than-before,const:something-before-and-after} to the
resulting set of linear inequalitites.

@@

Our approach proceedings in multiple steps:
\begin{inparaenum}[(i)]
\item Similarly as in \cite{generate-parikh-image}, we first describe
  each automaton as a network-flow problem, counting how often each
  transition is taken in a run of the automaton.
\item   We use a linear integer solver to discover that some
  paths through the automata are incompatible with
  \cref{const:more-inside-than-before,const:something-before-and-after}. This
  will allow us to discover an unreachable state, which allows further
  elimination. Finally, we will use case splitting to further
  subdivide the problem into bite-sized products of automata, both of
  which are empty, thus concluding that \cref{ex:string-constraints}
  is unsatisfiable.
\end{inparaenum}

\subsubsection{Eliminating paths through automata using flow analysis}\label{sec:a_1}

@@

\subsubsection{Using linear algebra to eliminate flow}\label{sec:intuition:algebra}

We then use linear arithmetic reasoning on register
equations~\eqref{eq:counter-sums} and flow
equations~\eqref{eq:flow-eqs} to simplify the counters associated
with transitions.
As an example, we start with $A'$. Substituting back
solutions to the equations for $A'$ to the automaton, we obtain the
automaton in \cref{fig:a_2}, where the notation~$a / t$ now denotes
a transition that accepts letter~$a$ and is taken~$t$ times.
In this case, the transitions of the automaton
can be specified purely in terms of the free
variables we care about representing:  the length of the automaton ($l_A$), the
start of the substring ($i$), and the length of the substring ($n$).

@@

Having obtained this representation, we conclude that $1 < n \leq 2$ from
\cref{const:something-before-and-after,const:something-before-and-after}. The
upper bound is obtained by the following reasoning. Since all transition
variables are nonnegative integers (a transition cannot be used a negative
amount of times), $n \leq 2$ from the $I_A$ to $F_A$ transition. Therefore, it
follows that $n=2$, which implies that the $I_A$ to $F_A$ transition can never
be used under these constraints and that we must use the two $\Char{d}$-labelled
transitions, both now taken $n-1 =1$ times. This immediately implies that the problem is
unsatisfiable, since we know this transition has no correspondance in automaton
$B$.

We repeat similar methods to arrive at the automata in \cref{fig:propagated}.

\begin{figure}[ht]
    \centering 
  \begin{subfigure}[b]{\autscale\textwidth}
    \centering
    \includegraphics[width=\textwidth]{a_annotated}
    \caption{ $A'$ with its associated transition variables in symbolic
    form.}\label{fig:aut_a_annotated}
      %\Description[SHORT]{LONG}
  \end{subfigure}\hfill%
  \begin{subfigure}[b]{\autscale\textwidth}
    \centering
    \includegraphics[width=\textwidth]{b_annotated}
    \caption{$B$ with its associated transition variables in symbolic forms.
    Note the large number of implicitly existence-quantified bound variables on
    transitions, suggesting that this automaton has a more complex structure
    with relation to its (free) target variable representing the string length
    than $A'$. The variables used here are fresh, and do not correspond to the
    ones previously used.}\label{fig:aut_b_annotated}
      %\Description[SHORT]{LONG}
  \end{subfigure}
  \caption{The automata after applying linear algebra and equality elimination
  to prune them.}\label{fig:propagated}
\end{figure}

@@

\subsubsection{Subsumption of connectivity constraints}\label{sec:intuition:subsume}
At this point we can conclude that none of the loops of $A'$ can become
disconnected since they are at the initial and final states, both of which are
reachable for any future assignments of the transition variables. This means
that all valid counter assignments of $A'$ are described by
\cref{eq:counter-sums} and the flow equations alone without the need for
expensive connectivity constraints.

\subsubsection{Case Splitting and propagation of connectivity constraints}\label{sec:intuition:split}
After concluding this, we have a choice of two paths through automaton $B$; the
upper through state $A$ or the lower through state $B$. We case split on a
problem-aware basis by selecting a transition variable that would disconnect
some strongly connected component of automaton $B$, in this case the transition
form $I_B$ to state $B$. We divide it into the cases used ($> 0$) or unused ($ =
0$) since those are the threshold values of the calculus, preferring the latter
option on the first-fail principle.

Having performed the split, we can conclude that the path from $I_B$ to state
$B$ is now unreachable without passing through a zero-times-used transition.
This means that the self-loop from state $B$ is also unusable, and thus must
also be used zero times.

\subsubsection{Computing products}\label{sec:intuition:materialise}
Now we have two small, stick-like automata after discounting all
transitions with a corresponding transition variable that is zero, and can try
to compute their product. Doing so, we will immediately notice that the
$\Char{d}$-labelled transition of automaton $A'$ has no correspondence in $B$,
leading to an empty product. We close the proof goal and backtrack with the
opposite constraint, choosing the upper $I_B$ to $A$ transition this time, the
only potentially productive choice we have left. However, proceding with
calculation of the product from the almost identical terms will be similarly
improductive, and we must conclude that the problem is unsatisfiable.

By putting off computing the product $B \times A'$ until after performing linear
reasoning on the number of times each transition must be taken and using that to
prune the terms, we have computed a smaller product than we would have with a
na\"ive approach. Moreover, since this automaton contains only one loop along
one main path, all runs through it can be described by the flow-preservation
equations described above without the need for costly connectivity-enforcing
constraints as in~\cite{generate-parikh-image}.

>>>>

%We will now proceed to show how we prove this using the three principles of
%\Calculus{}: problem-aware case splitting (sometimes called branching), lazy
%enforcement of connectivity constraints (essentially consisting of a constraint
%programming-style propagator), and lazy computation of products of automata.

An eager approach to finding the Parikh
image, as described in~\cite{generate-parikh-image}, would start by
computing the product $A' \times B$, translating it to a Presburger formula with
$i, n$, etc., as free variables, and then adding
\cref{const:more-inside-than-before,const:something-before-and-after} to the
resulting set of linear inequalitites. Scalability of this
construction is limited due to the size of the product automata to be computed,
and the complexity of the resulting Presburger formula.

@@

We will now proceed to show how our calculus~\Calculus{} can lazily
prove the unsatisfiability of the running example. The calculus
interleaves several reasoning principles, which we will later
define precisely as calculus rules:
\begin{inparaenum}[(i)]
\item Similarly as in \cite{generate-parikh-image}, we first describe
  each automaton as a flow network, counting how often each transition
  is taken in a run of the automaton. The flow constraints are an
  over-approximation of the possible accepting runs of an automaton.
\item We use a linear integer solver to simplify the flow constraints,
  and prune away paths through the automata that are not feasible.
\item We use a tailor-made propagation algorithm to identify disconnected
  transitions of the automata that can never be taken, and lazily add
  the corresponding constraints.
\item When propagation cannot infer further constraints, we use case
  splitting to sub-divide the problem into smaller parts.
\item Once the constituent automata of a product have become
  sufficiently small, we compute the precise product automaton.
\end{inparaenum}

\subsubsection{Approximating paths through automata using flow analysis}\label{sec:a_1}

@@

\subsubsection{Arithmetic flow simplification}\label{sec:intuition:algebra}

We then use linear arithmetic reasoning on register
equations~\eqref{eq:counter-sums} and flow
equations~\eqref{eq:flow-eqs} to simplify the counters associated
with transitions.
As an example, we start with $A'$. Substituting back
solutions to the equations for $A'$ to the automaton, we obtain the
automaton in \cref{fig:a_2}, where the notation~$a / t$ now denotes
a transition that accepts letter~$a$ and is taken~$t$ times.
In this case, the transitions of the automaton
can be specified purely in terms of the free
variables we care about representing:  the length of the accepted word ($l_A$), the
start of the substring ($i$), and the length of the substring ($n$).

@@

Having obtained this representation, we conclude that $1 < n \leq 2$
from
\cref{const:more-inside-than-before,const:something-before-and-after}. The
lower bound directly follows from
\cref{const:more-inside-than-before,const:something-before-and-after},
whereas the upper bound is obtained by the following reasoning. Since all
transition variables are non-negative integers (a transition cannot be
used a negative amount of times), $n \leq 2$ from the $I_A$ to $F_A$
transition. Therefore, it follows that $n=2$, which implies that the
$I_A$ to $F_A$ transition can never be used under these constraints
and that we must use the two $\Char{d}$-labelled transitions, both now
taken $n-1 =1$ times.
% This immediately implies that the problem is
%unsatisfiable, since we know this transition has no correspondance in
%automaton $B$.
%
Substituting the derived value of $n$, and applying similar
reasoning to automaton~$B$, we arrive at the simplified
automata in \cref{fig:propagated}.

\begin{figure}[ht]
    \centering 
  \begin{subfigure}{0.42\textwidth}
    \centering
    \includegraphics[width=\textwidth]{a_annotated}
    \vspace*{5ex}
    \caption{ $A'$ with its associated transition variables in symbolic
    form.}\label{fig:aut_a_annotated}
      %\Description[SHORT]{LONG}
    \vspace*{1.9ex}
  \end{subfigure}\hfill%
  \begin{subfigure}{0.54\textwidth}
    \centering
    \includegraphics[width=\textwidth]{b_annotated}
    \caption{$B$ with its associated transition variables in symbolic forms.
    The large number of implicitly existentially quantified
    variables~$x_i$ on
    transitions  suggests that this automaton has a more complex structure
    with relation to its (free) target variable representing the string length
    than $A'$. %The variables used here are fresh, and do not correspond to the
    % ones previously used.
  }\label{fig:aut_b_annotated}
      %\Description[SHORT]{LONG}
  \end{subfigure}
  \caption{The automata after arithmetic simplification.}\label{fig:propagated}
\end{figure}

@@

%\subsubsection{Connectivity reasoning}\label{sec:intuition:subsume}
%At this point we can conclude that none of the loops of $A'$ can become
%disconnected since they are at the initial and final states, both of which are
%reachable for any future assignments of the transition variables. This means
%that all valid counter assignments of $A'$ are described by
%\cref{eq:counter-sums} and the flow equations alone without the need for
%expensive connectivity constraints.

\subsubsection{Case Splitting and connectivity propagation}\label{sec:intuition:split}
We now have a choice of two paths through automaton $B$; the
upper through state $A$ or the lower through state $B$. Since
arithmetic reasoning and propagation are not able to resolve this choice,
we perform a case split by selecting a transition variable that would disconnect
some strongly connected component of automaton $B$, in this case the transition
from~$I_B$ to state $B$ guarded by variable~$x_{12}$.
 We split the reasoning into the cases $x_{12} > 0$ (transition used)
and $x_{12} = 0$ (transition unused). For presentation, we focus
on the latter case, as the former can be handled in a similar way.

In the case~$x_{12} = 0$ we can conclude
that the state~$B$ is now unreachable, which means that also
transitions outgoing from state $B$ are now unusable. Propagation can
therefore infer the additional equations $x_9 = 0$ and $x_{11} = 1$.

\subsubsection{Computing products}\label{sec:intuition:materialise}
After discounting all transitions that can no longer be taken, we are
left with two (small) flat automata, and can compute their product with
relative ease. By putting off computing the product $B \times A'$ until after
performing linear reasoning, and using that to prune transitions, we
have computed a smaller product than we would have with an eager
approach. Computing the product, we will immediately notice that the
$\Char{d}$-labelled transition of automaton $A'$ has no correspondence
in $B$, leading to an empty product. We can close the proof goal and
backtrack, and will eventually derive that the imposed constraints are
unsatisfiable.
